Liveness-Based Garbage Collection for Lazy Languages

### 2016-04-26 paper/paper.tex
Fixed WHNF reference issue

%% We use the  result of the analysis to annotate  each potential garbage
%% collection  point  in   the  program  with  a   set  of  deterministic
%% finite-state automata (DFA) describing the liveness at that point.
To  show the  effectiveness of  our method,  we implemented  a copying
collector that uses  the results of the liveness  analysis to preserve
live  objects,  both evaluated  (i.e.,  in  WHNF) and  closures.   Our
experiments confirm  that for  programs running with  a liveness-based
garbage  collector, there  is a  significant decrease  in peak  memory
requirements.   In addition,  a  sizable reduction  in  the number  of
collections  ensures that  in spite  of using  a more  complex garbage
collector, the execution  times of programs running  with liveness and
reachability-based collectors remain comparable.
\end{abstract}

@@

While  the runtime  system  of most  functional  languages includes  a
garbage collector to efficiently  reclaim memory, empirical studies on
Scheme~\cite{karkare06effectiveness}  and,  Haskell~\cite{rojemo96lag}
programs have shown that garbage  collectors leave uncollected a large
number of  memory objects that are  reachable but not live  (here {\em
  live} means the  object can potentially be used by  the program at a
later stage).  This results in unnecessary memory retention.

@@

Here  $\mathit{Loc}$ is  a countable  set  of locations  in the  heap.
A non-empty location either contains  a value in  \emph{WHNF} (a  number, the
empty  list  $\NIL$ or  a  \CONS\  cell  with  possibly  unevaluated
constituents) or  a {\em closure}.   A closure  is a pair  $\langle s,
\rho\rangle$ in  which $s$ is  an unevaluated application,  and $\rho$
maps free variables of $s$ to their respective locations. Since all data objects are boxed, we model an environment as a mapping
from the set  of variables of the program  $\mathit{Var}$ to locations
in the  heap.

>>>>

%% We use the  result of the analysis to annotate  each potential garbage
%% collection  point  in   the  program  with  a   set  of  deterministic
%% finite-state automata (DFA) describing the liveness at that point.
To  show the  effectiveness of  our method,  we implemented  a copying
collector that uses  the results of the liveness  analysis to preserve
live  objects,  both evaluated  and  closures.   Our
experiments confirm  that for  programs running with  a liveness-based
garbage  collector, there  is a  significant decrease  in peak  memory
requirements.   In addition,  a  sizable reduction  in  the number  of
collections  ensures that  in spite  of using  a more  complex garbage
collector, the execution  times of programs running  with liveness and
reachability-based collectors remain comparable.
\end{abstract}

@@

While  the runtime  system  of most  functional  languages includes  a
garbage collector to efficiently  reclaim memory, empirical studies on
Scheme~\cite{karkare06effectiveness}  and  Haskell~\cite{rojemo96lag}
programs have shown that garbage  collectors leave uncollected a large
number of  memory objects that are  reachable but not live  (here {\em
  live} means the  object can potentially be used by  the program at a
later stage).  This results in unnecessary memory retention.

@@

Here $\mathit{Loc}$  is a countable set  of locations in the  heap.  A
non-empty location either contains a {\em closure}, or a value in Weak
Head Normal Form (WHNF)\cite{Jones87}. In our case, a WHNF value
can be a number, the empty list  $\NIL$ or a \CONS\ cell with possibly
unevaluated   constituents.   A   closure  is   a  pair   $\langle  s,
\rho\rangle$ in  which $s$ is  an unevaluated application,  and $\rho$
maps free  variables of $s$  to their respective locations.  Since all
data objects are boxed, we model  an environment as a mapping from the
set of  variables of  the program $\mathit{Var}$  to locations  in the
heap.
